A Guide to Pre-Processing High-Throughput Animal Tracking Data

### 2021-02-16 05_calibration_data.Rmd
add breaker for patch emthod

@@

## Plot figure 7

>>>>

At this stage, users have successfully pre-processed their data from raw positions to residence patches.
Residence patches are essentially `sf` objects and can be visualised using the `sf` method for `plot`; for instance `plot(patch_sf_data)`.
Further sections reproduce the analyses in the main manuscript.

---

@@

## Main text Figure 6

### 2021-02-16 docs/ms_atlas_preproc_main_text.tex

To demonstrate the pipeline, we simulated a realistic movement track of 5,000 positions (unbiased correlated velocity model; UCVM) using the \texttt{R} package \texttt{smoove} \citep[][see Fig. 2.a]{gurarie2017}.
We added three kinds of error to the simulated track: (1) normally distributed small-scale offsets to the X and Y coordinates independently, (2) normally distributed large-scale offsets to a random subset (0.5 \%) of the positions, and (3) large-scale displacement of a continuous sequence of 300 of the 5,000 positions (indices 500 -- 800) (Fig. 2.a).
To demonstrate the residence patch method, we chose to simulate three independent rotational/advective correlated velocity movement (RACVM) tracks of 500 positions each ($\omega$ = 7, initial velocity = 0.1, $\mu$ = 0; see Gurarie et al. 2017), and connected them together with a roughly linear path (see Fig. 5.a).
RACVM models approximate the tracks of soaring birds which circle on thermals over a relatively small area, and move between thermals \citep[`thermalling'; ][]{gurarie2017, harel2016}.
This complex track structure provides a suitable challenge for the residence patch method and helps to demonstrate its generality.

@@

Sometimes, entire subsets of the track may be affected by the same large-scale location-error.
For instance, multiple consecutive positions may be roughly translated (geometrically) away from the real track and form `prolonged spikes', or `reflections' (see Fig. 2.b).
These cannot be corrected by targeted removal of individual positions, as in Bjorneraas and colleagues' approach (2010), since there are no positions with both high incoming and outgoing speeds.
Since filtering individual positions will not suffice, algorithms to correct such errors must take a track-level view, and target the displaced sequence overall.
Track-subset algorithms are likely to be system-specific, and may be challenging to conceptualise or implement.
In the case of prolonged spikes, one relatively simple solution is identifying the bounds of displaced segments, and removing positions between them.
% We show an illustrative example of such an algorithm in the form of track-subset filtering for prolonged spikes using the \texttt{atlastools} function \texttt{atl\_remove\_reflections} (Listing 5).
Users are strongly encouraged to visualise their data before and after applying such algorithms, 
We caution that these methods are not foolproof, and data that are heavily distorted by errors affecting entire track-subsets should be used with care when making further inferences.

@@

Users may question why thinning, which can obtain consensus positions over an interval and also reduce data-volumes should not be used directly on the raw data.
%%
We caution that thinning prior to excluding unrealistic movement and smoothing (Fig 4.b) can lead to preserving artefacts in the data, and estimates of essential metrics --- such as speed --- that are substantially different from the true value (Noonan et al. 2019; see Fig. 4.c).
In our example for instance, the data with errors included would have to be thinned to \sfrac{1}{30}\textsuperscript{th} of its volume for the median speed to be correctly estimated; an undesirable step if the aim is fine-scale tracking (Fig. 4.c).
The mis-estimation of track metrics could have knock-on consequences for the implementation of subsequent filters based on detecting unrealistic movement.
However, thinning before data-cleaning may have its place as a useful step before exploratory visualisation of the movement track, since reduced data-volumes are easier to handle for plotting software.
%
Thinning is implemented in \texttt{atlastools} using the \texttt{atl\_thin\_data} function, with either aggregation or resampling (specified by the \texttt{method} argument) over an interval using the \texttt{interval} argument.
% The column of timestamps must be named `time' and column classes except the identity columns (\textit{see below}) should allow averaging.
% The `resample' option returns a thinned dataset with all columns from the input data, but `aggregate' drops \texttt{COVXY}, as this cannot be propagated.
% Using `resample' returns the actual timestamp (in UNIX time) of each sample, while `aggregate' returns the mean timestamp (also in UNIX time).
% In both aggregation and resampling, an extra column \texttt{time\_agg} is added which has a uniform difference between each element corresponding to the user-defined thinning interval.
% Due to its development for ATLAS systems, the `aggregate' assumes the variance around coordinates is named \texttt{VARX} and \texttt{VARY}, and standard deviation around each position is named \texttt{SD}.
% These columns must be present together for the function to correctly handle the error.
% If there is no measure of error, the function simply returns the averaged position and covariates in each time interval.
Grouping variables' names (such as animal identity) may be passed as a character vector to the \texttt{id\_columns} argument (Listing 5).

>>>>

To demonstrate the pipeline, we simulated a realistic movement track of 5,000 positions (unbiased correlated velocity model; UCVM) using the \texttt{R} package \texttt{smoove} \citep[][see Fig. 2.a]{gurarie2017}.
We added three kinds of error to the simulated track: (1) normally distributed small-scale offsets to the X and Y coordinates independently, (2) normally distributed large-scale offsets to a random subset (0.5 \%) of the positions, and (3) large-scale displacement of a continuous sequence of 300 of the 5,000 positions (indices 500 -- 800) (Fig. 2.a).
To demonstrate the residence patch method, we chose to simulate three independent rotational-advective correlated velocity movement (RACVM) tracks of 500 positions each ($\omega$ = 7, initial velocity = 0.1, $\mu$ = 0; see Gurarie et al. 2017), and connected them together with a roughly linear path (see Fig. 5.a).
RACVM models approximate the tracks of soaring birds which circle on thermals over a relatively small area, and move between thermals \citep[`thermalling'; ][]{gurarie2017, harel2016}.
This complex track structure provides a suitable challenge for the residence patch method and helps to demonstrate its generality.

@@

Sometimes, entire subsets of the track may be affected by the same large-scale location-error.
For instance, multiple consecutive positions may be roughly translated (geometrically) away from the real track and form `prolonged spikes', or `reflections' (see Fig. 2.b).
These cannot be corrected by targeted removal of individual positions, as in Bjørneraas and colleagues' approach (2010), since there are no positions with both high incoming and outgoing speeds.
Since filtering individual positions will not suffice, algorithms to correct such errors must take a track-level view, and target the displaced sequence overall.
Track-subset algorithms are likely to be system-specific, and may be challenging to conceptualise or implement.
In the case of prolonged spikes, one relatively simple solution is identifying the bounds of displaced segments, and removing positions between them.
% We show an illustrative example of such an algorithm in the form of track-subset filtering for prolonged spikes using the \texttt{atlastools} function \texttt{atl\_remove\_reflections} (Listing 5).
Users are strongly encouraged to visualise their data before and after applying such algorithms, 
We caution that these methods are not foolproof, and data that are heavily distorted by errors affecting entire track-subsets should be used with care when making further inferences.

@@

Users may question why thinning, which can obtain consensus positions over an interval and also reduce data-volumes should not be used directly on the raw data.
%%
We caution that thinning prior to excluding unrealistic movement and smoothing (Fig 4.b) can lead to preserving artefacts in the data, and estimates of essential metrics --- such as speed --- that are substantially different from the true value \citep[see Fig. 4.c;][]{noonan2019}.
In our example for instance, the data with errors included would have to be thinned to \sfrac{1}{30}\textsuperscript{th} of its volume for the median speed to be correctly estimated; an undesirable step if the aim is fine-scale tracking (Fig. 4.c).
The mis-estimation of track metrics could have knock-on consequences for the implementation of subsequent filters based on detecting unrealistic movement.
However, thinning before data-cleaning may have its place as a useful step before exploratory visualisation of the movement track, since reduced data-volumes are easier to handle for plotting software.
%
Thinning is implemented in \texttt{atlastools} using the \texttt{atl\_thin\_data} function, with either aggregation or resampling (specified by the \texttt{method} argument) over an interval using the \texttt{interval} argument.
% The column of timestamps must be named `time' and column classes except the identity columns (\textit{see below}) should allow averaging.
% The `resample' option returns a thinned dataset with all columns from the input data, but `aggregate' drops \texttt{COVXY}, as this cannot be propagated.
% Using `resample' returns the actual timestamp (in UNIX time) of each sample, while `aggregate' returns the mean timestamp (also in UNIX time).
% In both aggregation and resampling, an extra column \texttt{time\_agg} is added which has a uniform difference between each element corresponding to the user-defined thinning interval.
% Due to its development for ATLAS systems, the `aggregate' assumes the variance around coordinates is named \texttt{VARX} and \texttt{VARY}, and standard deviation around each position is named \texttt{SD}.
% These columns must be present together for the function to correctly handle the error.
% If there is no measure of error, the function simply returns the averaged position and covariates in each time interval.
Grouping variable names (such as animal identity) may be passed as a character vector to the \texttt{id\_columns} argument (Listing 5).
